---
title: "Mini-Project #4"
author: "Jason Stasio"
format: 
  html:
    embed-resources: true
editor: visual
---

**Bayesian Analysis**

## Introduction

The goal of this project is to apply a Bayesian approach to modeling the probability that Rafael Nadal, one of the world's greatest men's clay-court tennis players, wins a point on his own serve against his primary rival, Novak Djokovic, during the French open tournament. It investigates how the data transforms three different scenarios/prior distributions, a noninfoirmative prior where we know nothing about the probability of Nadal scoring, an informative prior based on a previous clay-court match where Nadal won 46 out of 66 points with standard error 0.05657, and an informative prior based on a sports announcer's claim that Nadal wins the point about 75% of the time and never less than 70% of the time. First, the prior distributions are determined and graphed, with justification for choices included along with the code. Next, the data is applied to the prior to create the posterior distributions, which are determined and graphed. Calculations associated with these determinations are also included in the body using built-in latex. Along with graphing the posterior distributions, their means and associated 90% credible intervals are determined, with the method used to obtain these values given. Finally, a comparison and conclusion will be conducted to interpret the results.

## Body

#### Set-up

```{r}
# Clear the environment
rm(list = ls())

# Load in packages
library(tidyverse)
```

#### Prior Distributions

```{r}
ps <- seq(0, 1, length.out = 1000)

# Prior 1: Non-informative

# We will use a beta distribution with alpha=1 and beta=1 (equivalent to uniform(0,1)) to represent our non-informative prior since it is flat and thus any value between 0 and 1 holds equal density.

noninformative_alpha <- 1
noninformative_beta <- 1

noninformative_prior <- dbeta(ps, noninformative_alpha, noninformative_beta)

# Prior 2: Informative Based on a Clay-court Match

# Since we are given the number of points Nadal wins on his serve against Djokovic out of the total number of points for his serves, we have a proportion represented as a fraction that can take on values between 0 and 1. Thus, a beta distribution is a good choice for the prior. The general mean of the beta distribution can be set to the information given, a fraction, to solve for alpha and beta parameters. Let us note that we could also set it equal to the proportion represented as a decimal, solve for the general beta value depending on alpha, sequence alpha, then look for the values that give a standard error close to the one given. However, this would be unnecessary and computationally inefficient since we are given an exact value for the mean, which allows us to directly solve for specific alpha and beta parameters representative of the scenario.
```

$$ \frac{46}{66} = \frac{\alpha}{\alpha +\beta} \implies \alpha = 46,\text{ } \beta +\alpha = 66 \implies \beta = 66 - \alpha= 66 - 46=20$$

```{r}
ccm_informative_alpha <- 46
ccm_informative_beta <- 20

ccm_informative_prior <- dbeta(ps, ccm_informative_alpha, ccm_informative_beta)

# Prior 3: Informative Based on Sports Announcer

# Once again, we are given information pertaining to the proportion of points Nadal wins on his serves against Djokovic, so a beta distribution is a good choice for the prior. The sports announcer claims Nadal wins 75% of the points on his serve against Djokovic, which corresponds to the proportion 0.75. He also says he is "almost" sure Nadal wins no less than 70% of his serves, so we can adopt the strategy of solving for the beta parameter depending on alpha, sequencing through alpha values, calculating the distance to having a probability of 0.02 (a small value) of being less than 0.7 based on the parameters and filtering to find the smallest one. The parameters associated with the smallest distance to our target probability will be used. 
```

$$0.75 = \frac{\alpha}{\alpha + \beta}\implies 0.75\alpha +0.75\beta=\alpha\implies0.75\beta=0.25\alpha \implies \beta =\frac{0.25\alpha}{0.75} $$

```{r}
alphas <- seq(1, 1000, length.out = 5000)
betas <- (0.25 * alphas) / 0.75

target_prob <- 0.02
prob_less_2 <- pbeta(0.7, alphas, betas)

tibble(alphas, betas, prob_less_2) |>
  mutate(close_to_target = abs(prob_less_2 - target_prob)) |>
  filter(close_to_target == min(close_to_target))

sann_informative_alpha <- 252
sann_informative_beta <- 83.9

sann_informative_prior <- dbeta(ps, sann_informative_alpha, sann_informative_beta)

plot_df_prior <- tibble(ps, noninformative_prior,
                  ccm_informative_prior,
                  sann_informative_prior) |>
  pivot_longer(cols = -ps, names_to = "distribution", values_to = "density")

ggplot(data = plot_df_prior, aes(x = ps, y = density, colour = distribution)) +
  geom_line() +
  scale_colour_viridis_d(end = 0.9) +
  theme_minimal() +
  labs(x = "p")
```

#### Data

Since we are given data that has two outcomes, either Nadal scored the point on his serve or he did not, we can model it as a binomial distribution. We proved in class that combining a prior beta distribution with a binomial distribution will result in a posterior beta distribution with parameters:

$$\alpha_{post} = y_{obs} + \alpha, \text{ } \\ \beta_{post} = n - y_{obs} + \beta$$ Now, we can solve for the posterior distribution for each setting using the fact Nadal won 56 of the points from 84 possible points.

Non-informative: $$ \alpha_{post} = 56 + 1=57, \text{ } \\ \beta_{post} = 84 - 56+1=29$$

Informative based on clay-court match: $$\alpha_{post} = 56 + 46=102, \text{ } \\ \beta_{post} = 84 - 56+20=48$$

Informative based on sports announcer: $$\alpha_{post} = 56 + 252 = 308, \text{ } \\ \beta_{post} = 84 - 56 + 83.9=111.9$$

```{r}
# Plot the new posterior distributions
noninformative_posterior = dbeta(ps, 57, 29)
ccm_informative_posterior = dbeta(ps, 102, 48)
sann_informative_posterior = dbeta(ps, 308, 111.9)

plot_df_post <- tibble(ps, noninformative_posterior,
                  ccm_informative_posterior,
                  sann_informative_posterior) |>
  pivot_longer(cols = -ps, names_to = "distribution", values_to = "density")

ggplot(data = plot_df_post, aes(x = ps, y = density, colour = distribution)) +
  geom_line() +
  scale_colour_viridis_d(end = 0.9) +
  theme_minimal() +
  labs(x = "p")
```

Now, let us calculate the means of the posterior distributions.

Non-informative: $$\frac{\alpha_{post}}{\alpha_{post} + \beta_{post}} = \frac{57}{57+29}\approx 0.663$$

Informative based on clay-court match: $$\frac{\alpha_{post}}{\alpha_{post} + \beta_{post}} = \frac{102}{102+48} = 0.68 $$

Informative based on sports announcer: $$\frac{\alpha_{post}}{\alpha_{post} + \beta_{post}} = \frac{308}{308 + 111.9}\approx 0.734 $$

Furthermore, let us find 90% credible intervals for each of the scenarios.

Non-informative

```{r}
qbeta(c(0.05,0.95), 57, 29)

# According to our model, there is a 90% probability that the proportion of Nadal's serves where he wins the point is between 0.5772453 and 0.7440061
```

Informative based on clay-court match:

```{r}
qbeta(c(0.05,0.95), 102, 48)

# According to our model, there is a 90% probability that the proportion of Nadal's serves where he wins the point is between 0.6161904 and 0.7410715
```

Informative based on sports announcer

```{r}
qbeta(c(0.05,0.95), 308, 111.9)

# According to our model, there is a 90% probability that the proportion of Nadal's serves where he wins the point is between 0.6974328 and 0.7683171
```

## Comparison

Notice, the three posterior distributions are different from one another. This makes sense since they all follow beta distributions but with different parameters. In each setting, the beta parameter is smaller than the alpha parameter, but the distance between them varies. This is reflected by the fact the distributions have different variances, with smaller distances between parameters corresponding to larger variances. The posterior distribution that takes into account the previous clay-court match's data appears the most unchanged from its original prior distribution counterpart in terms of center, and it also has the second smallest variance. This suggests that this prior was in the closest alignment to the observed data, which would also be another explanation as to why it has the second smallest variance because the values are closer to the target compared to the noninformative prior. The prior and posterior distributions based on the sport's announcer's claims have the smallest variance, but this is likely due to bias in his reporting as we can see its center differs between the prior and posterior distributions to a greater degree compared to the other informative distribution setting. Thus, if a scenario had to be chosen, it would make sense to chose the prior distribution that takes into account the information given by the previous clay-court match data because it is the most unshifted from its center and has a relatively small variance that is not a result of bias. It also makes intuitive sense considering we have knowledge about previous matches, ruling out the need for a non-informative prior, and this informative prior takes into account data rather than opinion (which is more subject to bias).

## Conclusion

After analyzing the three scenarios, we found that the prior distribution based on the data from the previous clay-court match where Nadal won 46 out of 66 points with standard error 0.05657, is the best choice because it produces the posterior distribution with the second smallest variance and has the smallest shift from its center. A comparison between prior and posterior distributions shows that the sports announcer's setting is likely biased, but comes with the benefit of a smaller variance (e.g. bias-variance trade-off), while the other settings are unbiased and the informative prior has a less dramatic change than the non-informative, suggesting it is in closer alignment with the data. The mean of the posterior distribution for this setting is 0.68 and produces a 90% credible interval of (0.6161904 and 0.7410715). Thus, according to our model and Bayesian approach, there is a 90% probability that the proportion of Nadal's serves where he wins the point against his rival, Djokovic, during the French open tournament is between 0.6161904 and 0.7410715.
